Revisiting the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation using complex Empirical Orthogonal Functions

### 2021-08-11 analysis/paper/shceof.Rmd
Cambia coord_polar

geom_coords <- function() {
  list(
    no_grid,
    annotate("segment", 
             x = seq(0, 360 - 30, by = 30),
             xend =  seq(0, 360 - 30, by = 30),
             y = -90 + 15,
             yend = Inf,
             size = 0.1, alpha = 0.5),
    geom_hline(yintercept = seq(-90, 0, by = 15), size = 0.1, alpha = 0.5),
    shadowtext::geom_shadowtext(data = data.frame(x = 0, y = seq(-90 + 15, 0, 
                                                                 by = 15)),
                                aes(x, y, label = LatLabel(y)), size = 1.5, 
                                alpha = 0.7,
                                colour = "black",
                                bg.colour = "white")

@@

@@

In Section \@ref(methods) we describe the methods.
In Section \@ref(temporal) we describe the temporal variability and vertical coherence of the indices.
In Section \@ref(spatial), we analyse the spatial patterns of each complex EOF.
In Section \@ref(regressions) we study the spatial regressions with geopotential height and temperatuer. 
In Section \@ref(other-variables), we analyise the relationship between cEOF1 and Ozone, and between EOF2 and the Southern Annular Mode and the PSA. 
In Section \@ref(tropical) we look for tropical forcings that explain the variability of each cEOF. 
Finally, in Section \@ref(precipitation) we show the relationship between these modes of variabiltiy and precipitation in South America and Oceania.

@@

  geom_qmap(~.x[lat <= -20], size = 0.1) +
  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(10))  +
  scale_y_latitude(labels = NULL,
                   minor_breaks = NULL, 
                   limits = c(NA, -20)) +
  scale_x_longitude() +
  geom_coords() +
  coord_polar() +
  theme(axis.text = element_blank()) +
  facet_wrap(.~EOF, ncol = 2)  +
  tagger::tag_facets("rc")

@@

We performed all analysis in this paper using the R programming language [@rcoreteam2020], using data.table [@dowle2020] and metR [@campitelli2020] packages.
All graphics are made using ggplot2 [@wickham2009].
We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2019].

@@

  scale_fill_divergent_discretised(guide = guide_colorsteps_bottom()) +
  geom_qmap(~ .x[lat <= -20]) +
  scale_x_longitude(labels = NULL) +
  scale_y_latitude(labels = NULL,
                   limits = c(NA, -20),
                   minor_breaks = NULL) +
  scale_linetype_manual(values = c("1" = 1, "-1" = 2), labels = c("1" = "+", "-1" = "-"),
                        guide = "none") +
  facet_grid( lev  ~  cEOF,
              labeller = labeller(lev = lab_lev,
                                  cEOF = var)) +
  coord_quickmap(ylim = c(NA, -20)) +
  theme(legend.title = element_blank()) +
  coord_polar() +
  geom_coords() +
  tag_facets(tag = "cr")

@@

    geom_qmap(~ .x[lat <= -20]) +

    scale_x_longitude(labels = NULL) +
    scale_y_latitude(labels = NULL,
                     limits = c(NA, -20),
                     minor_breaks = NULL) +
    grid_y +
    facet +
    coord_polar() +
    geom_coords() +
    tag_facets("rc")
}
```

@@

eof_filtered %>%
  .$left %>%
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
  scale_fill_divergent(guide = "none") +
  geom_qmap(~.x[lat <= -20]) +
  scale_x_longitude() +
  scale_y_latitude() +
  coord_polar() +
  geom_coords() +
  theme(axis.text = element_blank())

@@

  scale_fill_divergent_discretised("Air temperature", 
                                   # limits = c(-2.25, 2.25),
                                   guide = guide_colorsteps_bottom(15)) + 
  scale_y_level() +
  scale_linetype(guide = "none") +
  scale_x_longitude(ticks = 60) +
  facet_grid(~term) +
  axis_labs_smol +
  coord_cartesian(clip = "off") +
  tag_facets()
```

@@

  geom_contour_pval(aes(z = p.value)) +
  geom_hline(yintercept = o3wave_lats, size = 0.3, alpha = 0.8) +
  geom_qmap(~.x[lat <= -20]) +
  scale_fill_divergent_discretised("Total Ozone Column", 
                                   limits = c(-30, 30),  
                                   guide = guide_colorsteps_bottom(15)) + 
  scale_x_longitude(labels = NULL) +
  scale_y_latitude(labels = NULL,
                   limits = c(NA, -20),
                   minor_breaks = NULL) +
  scale_linetype(guide = "none") +
  coord_polar() +
  geom_coords() +
  facet_grid(cEOF~term) +
  tag_facets("panel")

@@

    geom_contour_fill(data = sst_gdata,
                      aes(z = estimate, fill = ..level..), 
                      breaks = AnchorBreaks(exclude = 0)) +
    geom_contour_pval(data = sst_gdata, aes(z = p_val)) +
    scale_fill_divergent_discretised("SST", 
                                     guide = guide_colorsteps_bottom(order = 2)) +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_mag(guide = "none") +
    facet_grid(term ~ var) +
    coord_quickmap(ylim = c(NA, 10)) +
    theme(legend.box = "vertical") +
    tag_facets("cr", position = "bl") 
}
```

@@

add_continent <- function(dt) {
  dt[, continent := fcase(lon %between% continents$africa$lon &
                            lat %between% continents$africa$lat, "africa",
                          lon %between% continents$oceania$lon &
                            lat %between% continents$oceania$lat, "oceania",
                          lon %between% continents$america$lon &
                            lat %between% continents$america$lat, "america")] %>%
    .[, continent := factor(continent, levels = c("africa", "oceania", "america"), ordered = TRUE)] %>%
    .[]
}
```

@@

    # scale_fill_divergent() +
    scale_fill_divergent_discretised(NULL, 
                                     # limits = c(-.8, .8),
                                     # oob = scales::squish,
                                     low = "#8E521C",
                                     high = "#00665E",
                                     guide = guide_colorsteps_bottom(15)) +
    # scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(15)) +
    geom_contour_pval(aes(z = p.value)) +
    geom_qmap(~.x %>% .[lat %between% c(-60, max(data$lat))] %>% 
                add_continent() %>% filter_continent(),
              keep = 1) +

    scale_x_longitude(ticks = 15) +
    scale_y_latitude(ticks = 15) +
    ggh4x::facet_nested(term ~ cEOF) +
    axis_labs_smol +
    tag_facets("cr") +
    coord_quickmap()
}
```

```{r sesa}
sesa <- list(lon = c(300, 315),
             lat = c(-35, -22))
```

@@

```{r box_australia}
box_australia <- data.frame(cEOF = "cEOF2", 
                            lon = c(130, 155),
                            lat = c(-40, -15))
```

>>>>

geom_coords <- function() {

  list(
    no_grid,
    annotate("segment",
             y = seq(-90, 0, by = 15),
             yend = seq(-90, 0, by = 15),
             x = 0,
             xend = 360,
             size = 0.1, alpha = 0.5),

    annotate("segment",
             x = seq(0, 360 - 30, by = 30),
             xend =  seq(0, 360 - 30, by = 30),
             y = -90 + 15,
             yend = Inf,
             size = 0.1, alpha = 0.5),
    shadowtext::geom_shadowtext(data = data.frame(x = 0, y = seq(-90 + 15, 0, 
                                                                 by = 15)),
                                aes(x, y, label = LatLabel(y)), size = 1.5, 
                                alpha = 0.7,
                                colour = "black",
                                bg.colour = "white")

@@

coord_polar <- function(ymax = -20) {

  x <- c(seq(0, 360, length.out = 40), 
         seq(360, 0, length.out = 40), 
         0)
  y <- c(rep(ymax, length.out = 40), 
         rep(60, length.out = 40), 
         ymax)

  cbind(x, y) |> 
    list() |> 
    sf::st_polygon() |> 
    sf::st_sfc(crs = "+proj=latlong") -> white

  list(
    geom_sf(data = white, inherit.aes = FALSE, 
            fill = "white", 
            colour = "white", size = 2),
    coord_sf(ylim = c(-90, ymax), 
             lims_method = "box",
             crs = "+proj=laea +lat_0=-90",
             default_crs = "+proj=longlat",
             label_axes =  "----")
  )
}

@@

In Section \@ref(methods) we describe the methods.
In Section \@ref(spatial), we analyse the spatial patterns of each complex EOF.
In Section \@ref(regressions) we study the spatial regressions with geopotential height and temperatuer. 
In Section \@ref(other-variables), we analyise the relationship between cEOF1 and Ozone, and between EOF2 and the Southern Annular Mode and the PSA. 
In Section \@ref(tropical) we look for tropical forcings that explain the variability of each cEOF. 
Finally, in Section \@ref(precipitation) we show the relationship between these modes of variabiltiy and precipitation in South America and Oceania.

@@

  geom_qmap(size = 0.1) +
  scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(10))  +
  scale_y_latitude(labels = NULL,
                   minor_breaks = NULL, 
                   limits = c(NA, -20)) +
  scale_x_longitude() +
  geom_coords() +
  coord_polar() +
  theme(axis.text = element_blank()) +
  facet_wrap(.~EOF, ncol = 2)  +
  tagger::tag_facets("rc")

@@

We performed all analysis in this paper using the R programming language [@rcoreteam2020], using data.table [@dowle2020] and metR [@campitelli2020] packages.
All graphics are made using ggplot2 [@wickham2009].
We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2020].

@@

  scale_fill_divergent_discretised(guide = guide_colorsteps_bottom()) +
  geom_qmap() +
  scale_x_longitude(labels = NULL) +
  scale_y_latitude(labels = NULL,
                   minor_breaks = NULL) +
  scale_linetype_manual(values = c("1" = 1, "-1" = 2),
                        labels = c("1" = "+", "-1" = "-"),
                        guide = "none") +
  facet_grid( lev  ~  cEOF,
              labeller = labeller(lev = lab_lev,
                                  cEOF = var)) +
  theme(legend.title = element_blank()) +
  geom_coords() +
  coord_polar() +

  tag_facets(tag = "cr")

@@

    geom_qmap() +

    scale_x_longitude(labels = NULL) +
    scale_y_latitude(labels = NULL,
                     minor_breaks = NULL) +
    grid_y +
    facet +
    geom_coords() +
    coord_polar() +
    tag_facets("rc")
}
```

@@

eof_filtered %>%
  .$left %>%
  periodic(lon = c(0, 360)) %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
  geom_contour_tanaka(aes(z = eof), breaks = AnchorBreaks(0, exclude = 0)) +
  scale_fill_divergent(guide = "none") +
  geom_qmap() +
  scale_x_longitude() +
  scale_y_latitude() +
  geom_coords() +
  coord_polar() +

  theme(axis.text = element_blank())

@@

  scale_fill_divergent_discretised("Air temperature", 
                                   # limits = c(-2.25, 2.25),
                                   guide = guide_colorsteps_bottom(15)) + 
  scale_y_level(minor_breaks= NULL, 
                sec.axis = sa_height_axis(breaks = seq(0, 50, by = 10),
                                          labels = function(x) paste0("~", x, " km"))) +
  scale_linetype(guide = "none") +
  scale_x_longitude(ticks = 60) +
  facet_grid(~term) +
  axis_labs_smol +
  coord_cartesian(clip = "off") +
  tag_facets()
```

@@

  geom_contour_pval(aes(z = p.value)) +
  annotate("segment", 
           y  = o3wave_lats,
           yend = o3wave_lats,
           x = 0,
           xend = 360,
           size = 0.3, alpha = 0.8
  ) +
  geom_qmap() +
  scale_fill_divergent_discretised("Total Ozone Column", 
                                   limits = c(-30, 30),  
                                   guide = guide_colorsteps_bottom(15)) + 
  scale_x_longitude(labels = NULL) +
  scale_y_latitude(labels = NULL,
                   minor_breaks = NULL) +
  scale_linetype(guide = "none") +
  geom_coords() +
  coord_polar() +
  facet_grid(cEOF~term) +
  tag_facets("panel")

@@

    geom_contour_fill(data = sst_gdata,
                      aes(z = estimate, fill = ..level..), 
                      breaks = AnchorBreaks(exclude = 0)) +
    geom_contour_pval(data = sst_gdata, aes(z = p_val)) +
    scale_fill_divergent_discretised("SST", 
                                     guide = guide_colorsteps_bottom(order = 2)) +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_mag(guide = "none") +
    facet_grid(term ~ var) +
    # coord_quickmap(ylim = c(NA, 10)) +
    coord_sf(ylim = c(-90, 0),
             crs = "+proj=longlat +pm=180 +over",
             default_crs = "+proj=longlat +over") + 
  theme(legend.box = "vertical") +
    tag_facets("cr", position = "bl") 
}
```

@@

add_continent <- function(dt) {
  dt[, continent := fcase(lon %between% continents$africa$lon &
                            lat %between% continents$africa$lat, "africa",
                          lon %between% continents$oceania$lon &
                            lat %between% continents$oceania$lat, "oceania",
                          lon %between% continents$america$lon &
                            lat %between% continents$america$lat, "america")] %>%
    .[, continent := factor(continent, levels = c("africa", "oceania", "america"), ordered = TRUE)] %>%
    .[]
}

@@

    # scale_fill_divergent() +
    scale_fill_divergent_discretised(NULL, 
                                     # limits = c(-.8, .8),
                                     # oob = scales::squish,
                                     low = "#8E521C",
                                     high = "#00665E",
                                     guide = guide_colorsteps_bottom(15)) +
    # scale_fill_divergent_discretised(NULL, guide = guide_colorsteps_bottom(15)) +
    geom_contour_pval(aes(z = p.value)) +
    geom_qmap(keep = 1) +

    scale_x_longitude(ticks = 15) +
    scale_y_latitude(ticks = 15) +
    ggh4x::facet_nested(term ~ cEOF) +
    axis_labs_smol +
    tag_facets("cr") +
      coord_sf(ylim = continents[[which_continent]]$lat,
               xlim = continents[[which_continent]]$lon,
               crs = "+proj=longlat +pm=180 +over",
               default_crs = "+proj=longlat +over")

}
```

```{r boxes}
sesa <- list(lon = c(300, 315),
             lat = c(-35, -22))
box_australia <- data.frame(cEOF = "cEOF2", 
                            lon = c(130, 155),
                            lat = c(-40, -15))
```

@@
